---
title: "The auto selector and racing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The auto selector and racing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(optimix)
```

## Why per-instance selection

No optimiser wins everywhere. Rice framed this in 1976 as the
algorithm-selection problem: pick, per problem instance, the portfolio
member that will perform best. `method = "auto"` is the optimix answer.
It reads what the problem declares (design space, noise, expense, budget),
commits to an engine, and records the decision in the result.

The examples below use a shifted multimodal test function, a Rastrigin-style
surface whose global minimum sits at 1.3 in each coordinate:

```{r objective}
shifted <- function(x) {
  z <- x - 1.3
  sum(z^2 - 10 * cos(2 * pi * z) + 10)
}
```

## Tier 1: the rules

When the problem's declaration already determines a good choice, `auto`
applies a transparent rule and spends no evaluations on selection. A noisy
objective goes to a noise-tolerant global engine. An expensive objective goes
to Bayesian optimisation, which spends few true evaluations. A permutation
design space goes to the native simulated annealer, described in
*Combinatorial optimisation and delta evaluation*. And a small budget stays
with the rules, because probing the landscape would eat too much of it. A
noisy objective also never races -- ranking engines on one short noisy trial
each would rank them on luck -- so its rule commits directly:

```{r tier1-noisy, eval = requireNamespace("GenSA", quietly = TRUE)}
noisy_prob <- optim_problem(
  fn = function(x) sum(x^2) + rnorm(1, sd = 0.05),
  space = space_box(lower = c(-3, -3), upper = c(3, 3)),
  objective = objective_noisy(sd = 0.05),
  max_evals = 500, seed = 42
)
optimix(noisy_prob)$provenance[c("engine", "why", "tier")]
```

## Tier 3: racing

For a deterministic objective with an ample budget, `auto` escalates from
rules to evidence. It samples the landscape once with a Latin-hypercube
design, gives each installed candidate engine a short trial from the best
sampled point as a shared warm start, and commits the remaining budget to the
trial leader, warm-started from the leader's own best. The landscape sample
and every trial evaluation are charged against the budget, and the final run
receives only what is left.

```{r auto-race, eval = requireNamespace("GenSA", quietly = TRUE) && requireNamespace("cmaes", quietly = TRUE)}
set.seed(1)
res <- optimix(shifted, lower = c(-5, -5), upper = c(5, 5))
res
res$provenance[c("engine", "tier", "why")]
res$provenance$race
```

The provenance is how a result explains itself: the engine that ran
(`engine`), why it was chosen (`why`), and which tier decided (`tier` is 1
for a rule and 3 for a race). A raced result also keeps the trial scores.

## Asking for a race directly

`method = "race"` skips the deliberation and races the installed global
engines immediately. It needs a continuous box space and at least two
installed candidates:

```{r explicit-race, eval = requireNamespace("GenSA", quietly = TRUE) && requireNamespace("cmaes", quietly = TRUE)}
set.seed(2)
raced <- optimix(shifted, lower = c(-5, -5), upper = c(5, 5),
                 method = "race")
raced$provenance$why
```

## Budgets are respected

`max_evals` caps the true objective evaluations -- a soft cap, since an
engine may finish the internal step it is in -- and the reported count is
the honest total across every stage of the search. A budget of 200 is too
small for a race here (the landscape sample alone would cost a quarter of
it), so `auto` stays with the tier-1 rule; when the chunk below executes,
`capped$counts[["function"]]` sits at or just above 200, confirming the cap
was honoured rather than silently overrun:

```{r budget-cap, eval = requireNamespace("GenSA", quietly = TRUE)}
set.seed(3)
capped <- optimix(shifted, lower = c(-5, -5), upper = c(5, 5),
                  max_evals = 200)
capped$provenance$tier
capped$counts[["function"]]
```

## The evidence

The comparison is against `zakharov`, `salomon`, and `bohachevsky` -- three
functions held out from the rules `auto` was designed against -- each run at
3 dimensions x 6 seeds x 2 budgets (n = 54 instances per budget). On this
held-out set `auto` beat the best fixed single engine (`gensa`) where it
matters most: at a constrained budget of 300 evaluations per dimension it
solved more instances (success 0.63 against 0.56) at better median quality
and lower cost. At a generous budget of 1000 evaluations per dimension it
matched the single-best engine's quality (an equal-or-tighter median gap)
with about forty per cent fewer evaluations, but its success rate trails
single-best's at that budget (0.685 against 0.759) -- racing spends part of
the budget on trials that a strong single global would have spent solving,
which costs a little when nearly every instance already favours that one
engine. The efficiency win is real; the generous-budget success rate is the
honest price of it.

A feature-based shortcut -- route problems that look smooth to a cheap local
solver -- was prototyped and rejected on the same held-out evidence: one
function is multimodal yet looks perfectly quadratic to a coarse sample (a
bowl hiding a cosine ripple), so no feature threshold routes it safely.
Racing is feature-free, resolving such cases by actually trying the engines,
and that is why the escalation is a race rather than a landscape heuristic.
