---
title: "The auto selector and racing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The auto selector and racing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~2 s under rmarkdown::render() with GenSA, cmaes and
     nloptr all installed (the race-sweep figure re-runs three engines
     across six budgets); a build missing one of the three still renders
     every section, with the gated ones replaced by a skip sentence.
     macOS arm64, R 4.5.1, one core. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150
)
```

```{r library}
library(optimix)
```

## Why

An agronomist running the same nitrogen-response cost surface across dozens
of trial sites does not want to hand-pick a different optimiser for each
site, and she does not want to trust a single fixed choice either, because
no optimiser wins everywhere. Rice framed this in 1976 as the
algorithm-selection problem: pick, per problem instance, the portfolio
member that will perform best. `method = "auto"` is the optimix answer. It
reads what the problem declares (design space, noise, expense, budget),
commits to an engine, and records the decision in the result, so the
agronomist gets one call that works out, site by site, which optimiser to
trust, and a written reason she can check.

## What

`.auto_select()` (reached through `method = "auto"`, optimix's default)
runs in two tiers. Tier 1 is a transparent rule read straight off what the
problem declares -- no extra evaluations spent. Tier 3 escalates to racing:
a short trial for each installed candidate engine, then the remaining
budget committed to the trial leader. Every result carries a `provenance`
list recording the `engine` that ran, `why` it was chosen, and which `tier`
decided; a raced result additionally keeps the trial scores in
`provenance$race`. `method = "race"` reaches the same racing machinery
directly, skipping the tier-1 deliberation.

## Do

### The objective

The examples below use a shifted multimodal test function, a synthetic
Rastrigin-style surface whose global minimum sits at 1.3 in each
coordinate:

```{r objective}
shifted <- function(x) {
  z <- x - 1.3
  sum(z^2 - 10 * cos(2 * pi * z) + 10)
}
```

### Engines available on this build

Three sections below need an optional engine each, or a pair of them:
`GenSA` for the noisy-objective rule and every race, `cmaes` alongside it
for racing specifically, and `nloptr` (for `nloptr_bobyqa`) for the
budget-sweep figure. Each guarded chunk checks for what it needs and prints
one sentence in its place when an engine is missing.

```{r engines}
has_gensa <- requireNamespace("GenSA", quietly = TRUE)
has_cmaes <- requireNamespace("cmaes", quietly = TRUE)
has_nloptr <- requireNamespace("nloptr", quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
race_ok <- has_gensa && has_cmaes
sweep_ok <- race_ok && has_nloptr && has_ggplot2
```

### Tier 1: the rules

When the problem's declaration already determines a good choice, `auto`
applies a transparent rule and spends no evaluations on selection. A noisy
objective goes to a noise-tolerant global engine. An expensive objective
goes to Bayesian optimisation, which spends few true evaluations. A
permutation design space goes to the native simulated annealer, described
in *Combinatorial optimisation and delta evaluation*. And a small budget
stays with the rules, because probing the landscape would eat too much of
it. A noisy objective also never races -- ranking engines on one short
noisy trial each would rank them on luck -- so its rule commits directly:

```{r tier1-noisy, eval = has_gensa, echo = has_gensa}
noisy_prob <- optim_problem(
  fn = function(x) sum(x^2) + rnorm(1, sd = 0.05),
  space = space_box(lower = c(-3, -3), upper = c(3, 3)),
  objective = objective_noisy(sd = 0.05),
  max_evals = 500, seed = 42
)
res_noisy <- optimix(noisy_prob)
res_noisy$provenance[c("engine", "why", "tier")]
```

```{r tier1-noisy-skip, eval = !has_gensa, echo = FALSE, results = "asis"}
cat("GenSA is not installed, so the noisy-objective rule is not run here;",
    "the rule itself (route to a noise-tolerant global engine, never a",
    "race) does not depend on which engine ends up installed.\n")
```

### Tier 3: racing

For a deterministic objective with an ample budget, `auto` escalates from
rules to evidence. It samples the landscape once with a Latin-hypercube
design, gives each installed candidate engine a short trial from the best
sampled point as a shared warm start, and commits the remaining budget to
the trial leader, warm-started from the leader's own best. The landscape
sample and every trial evaluation are charged against the budget, and the
final run receives only what is left.

```{r auto-race, eval = race_ok, echo = race_ok}
set.seed(1)
res <- optimix(shifted, lower = c(-5, -5), upper = c(5, 5))
res
res$provenance[c("engine", "tier", "why")]
res$provenance$race
```

```{r auto-race-skip, eval = !race_ok, echo = FALSE, results = "asis"}
cat("Racing needs both GenSA and cmaes installed; at least one is missing",
    "here, so this race is not run.\n")
```

The provenance is how a result explains itself: the engine that ran
(`engine`), why it was chosen (`why`), and which tier decided (`tier` is 1
for a rule and 3 for a race). A raced result also keeps the trial scores.

### Asking for a race directly

`method = "race"` skips the deliberation and races the installed global
engines immediately. It needs a continuous box space and at least two
installed candidates:

```{r explicit-race, eval = race_ok, echo = race_ok}
set.seed(2)
raced <- optimix(shifted, lower = c(-5, -5), upper = c(5, 5),
                 method = "race")
raced$provenance$why
```

```{r explicit-race-skip, eval = !race_ok, echo = FALSE, results = "asis"}
cat("The same two engines are needed for an explicit race; this chunk is",
    "not run without both installed.\n")
```

### Budgets are respected

`max_evals` caps the true objective evaluations -- a soft cap, since an
engine may finish the internal step it is in -- and the reported count is
the true total across every stage of the search. A budget of 200 is too
small for a race here (the landscape sample alone would cost a quarter of
it), so `auto` stays with the tier-1 rule; when the chunk below executes,
`capped$counts[["function"]]` sits at or just above 200, confirming the cap
was honoured rather than silently overrun:

```{r budget-cap, eval = has_gensa, echo = has_gensa}
set.seed(3)
capped <- optimix(shifted, lower = c(-5, -5), upper = c(5, 5),
                  max_evals = 200)
capped$provenance$tier
capped$counts[["function"]]
```

```{r budget-cap-skip, eval = !has_gensa, echo = FALSE, results = "asis"}
cat("This budget-cap check needs GenSA installed; it is not run here.\n")
```

### Why the race commits to a leader

A single race, run once, shows which engine won; it does not show how the
three race candidates behave as the budget grows, which is the reason the
tier-3 rule commits the *remaining* budget to the trial leader rather than
splitting it evenly. The sweep below re-solves the same shifted surface
with each of `gensa`, `cmaes`, and `nloptr_bobyqa` directly, at six budgets
from 30 to 1000 evaluations, and records the best value each reaches:

```{r race-sweep, eval = sweep_ok, echo = sweep_ok}
sweep_budgets <- c(30L, 60L, 120L, 250L, 500L, 1000L)
sweep_engines <- c("gensa", "cmaes", "nloptr_bobyqa")
sweep_grid <- expand.grid(
  engine = sweep_engines, budget = sweep_budgets, stringsAsFactors = FALSE
)
s1 <- nrow(sweep_grid)
sweep_rows <- vector("list", s1)
for (i1 in seq_len(s1)) {
  prob_i <- optim_problem(
    fn = shifted, space = space_box(lower = c(-5, -5), upper = c(5, 5)),
    max_evals = sweep_grid$budget[i1], seed = 5
  )
  res_i <- optimix(prob_i, method = sweep_grid$engine[i1])
  sweep_rows[[i1]] <- data.frame(
    engine = sweep_grid$engine[i1], budget = sweep_grid$budget[i1],
    value = res_i$value, evals = res_i$counts[["function"]]
  )
} # ends i1, over the engine x budget grid
sweep_df <- do.call(rbind, sweep_rows)
sweep_df
```

```{r race-sweep-kable, echo = FALSE, eval = sweep_ok}
final_df <- sweep_df[sweep_df$budget == max(sweep_budgets), ]
knitr::kable(
  final_df[, c("engine", "budget", "value", "evals")],
  digits = 4, row.names = FALSE,
  caption = paste0(
    "Best value each engine reaches by the largest budget in the sweep (",
    max(sweep_budgets), " evaluations)."
  )
)
```

```{r fig-race-sweep, echo = sweep_ok, eval = sweep_ok, fig.cap = "Best value found against the true evaluation budget, for the three engines the tier-3 race chooses between. nloptr_bobyqa (a local search) settles early and stops improving regardless of budget; cmaes and gensa keep improving as the budget grows, with gensa reaching the global optimum by budget 500. This is what a single race's trial-then-commit step is racing for."}
ggplot2::ggplot(
  sweep_df, ggplot2::aes(x = evals, y = value, colour = engine)
) +
  ggplot2::geom_line() +
  ggplot2::geom_point(size = 2) +
  ggplot2::scale_colour_viridis_d(name = "engine") +
  ggplot2::labs(
    x = "true objective evaluations",
    y = "best value found",
    title = "Convergence of the three raced engines, by budget"
  ) +
  ggplot2::theme_minimal()
```

```{r race-sweep-skip, eval = !sweep_ok, echo = FALSE, results = "asis"}
cat("The budget sweep needs GenSA, cmaes and nloptr all installed, plus",
    "ggplot2 for the figure; at least one is missing here, so this",
    "section is not run.\n")
```

### The evidence

The comparison is against `zakharov`, `salomon`, and `bohachevsky` -- three
functions held out from the rules `auto` was designed against -- each run
at 3 dimensions x 6 seeds x 2 budgets (n = 54 instances per budget). These
figures come from the package's own tier-2/tier-3 held-out validation
study and are not recomputed in this vignette. On that held-out set `auto`
beat the best fixed single engine (`gensa`) where it matters most: at a
constrained budget of 300 evaluations per dimension it solved more
instances (success 0.63 against 0.56) at better median quality and lower
cost. At a generous budget of 1000 evaluations per dimension it matched the
single-best engine's quality (an equal-or-tighter median gap) with about
forty per cent fewer evaluations, but its success rate trails single-best's
at that budget (0.685 against 0.759) -- racing spends part of the budget on
trials that a strong single global would have spent solving, which costs a
little when nearly every instance already favours that one engine. The
efficiency win is real; the generous-budget success rate is the price
paid for it.

A feature-based shortcut -- route problems that look smooth to a cheap
local solver -- was prototyped and rejected on the same held-out evidence:
one function is multimodal yet looks perfectly quadratic to a coarse
sample (a bowl hiding a cosine ripple), so no feature threshold routes it
safely. Racing is feature-free, resolving such cases by actually trying
the engines, and that is why the escalation is a race rather than a
landscape heuristic.

## Read

```{r read-text, echo = FALSE, results = "asis"}
if (has_gensa) {
  cat(
    "The noisy-objective rule commits to",
    paste0("`", res_noisy$provenance$engine, "`"),
    "at tier", res_noisy$provenance$tier, "with no race, matching the",
    "stated rule: a noisy objective is never raced.\n\n"
  )
} else {
  cat(
    "GenSA is not installed on this build, so the noisy-objective rule is",
    "not exercised above; the paragraphs below cover only what did run.\n\n"
  )
}
if (race_ok) {
  trial_score <- round(unname(res$provenance$race[res$provenance$engine]), 6)
  final_score <- round(res$value, 6)
  commit_sentence <- if (trial_score - final_score > 1e-6) {
    paste(
      "the remainder of the budget bought a materially better answer",
      "than the trial alone found."
    )
  } else {
    paste(
      "on this instance the short trial alone had already located the",
      "exact optimum, so the remaining budget confirmed rather than",
      "improved that answer -- not every race needs its commit step to",
      "do further work."
    )
  }
  cat(
    "On the deterministic shifted surface, the tier-3 race commits to",
    paste0("`", res$provenance$engine, "`"),
    paste0("(tier ", res$provenance$tier, "),"),
    "reaching a value of", final_score, "against a trial score",
    "of", trial_score, "for that same engine's own short trial --",
    commit_sentence,
    "The explicit race (`method = \"race\"`) reasons the same way:",
    paste0("`", raced$provenance$why, "`."),
    "At a budget of 200, too small for a race, `auto` stays with the",
    paste0("tier-1 rule (tier ", capped$provenance$tier, "), and"),
    capped$counts[["function"]], "evaluations were spent against that cap",
    "-- at or just above 200, confirming the budget was honoured rather",
    "than silently overrun.\n\n"
  )
} else {
  cat(
    "GenSA and cmaes are not both installed on this build, so no race",
    "runs above; the tier-3 mechanics described in *What* are unexercised",
    "here.\n\n"
  )
}
if (sweep_ok) {
  fastest_to_global <- sweep_df[
    sweep_df$engine == "gensa" & sweep_df$value < 1e-6, "budget"
  ]
  bobyqa_evals <- unique(
    sweep_df[sweep_df$engine == "nloptr_bobyqa", "evals"]
  )
  cat(
    "The sweep makes the reason for that commit-to-the-leader design",
    "visible: `nloptr_bobyqa` stops improving once its own local search",
    "declares convergence, spending only",
    paste(sort(bobyqa_evals), collapse = "-"),
    "true evaluations regardless of how large a budget it is offered,",
    "and settling on a value well short of the global optimum. `cmaes`",
    "and `gensa` both keep using every evaluation they are given and",
    "keep improving; `gensa` is the first to reach the global optimum,",
    "at a budget of", min(fastest_to_global), "evaluations. A single",
    "short trial (the race's own first step) could not have shown this",
    "-- only a sweep across budgets does -- which is exactly why the race",
    "commits the remaining budget to whichever engine's short trial led,",
    "rather than trusting the trial's ranking to hold at every budget.\n"
  )
} else {
  cat(
    "The budget-sweep figure needs GenSA, cmaes, nloptr and ggplot2 all",
    "installed; at least one is missing here, so the paragraph above is",
    "the only account of that comparison on this build.\n"
  )
}
```

## Limits

The sweep above re-solves one synthetic surface at one seed; it shows the
shape of three engines' convergence on that instance, not a general
ranking, and the held-out study cited in *The evidence* is what actually
supports the claim that `auto` beats a fixed single engine on average. The
race itself commits on the strength of a single short trial per candidate,
which is why `.auto_select()` never races a noisy objective -- a
single-trial ranking on noise would rank engines on luck rather than
performance, and nothing in this vignette exercises that failure mode
directly. Every example here is a continuous box in two or three
dimensions; permutation spaces route around racing entirely, to the native
annealer covered in *Combinatorial optimisation and delta evaluation*.

## What to read next

*Getting started with optimix* covers the two front doors and the result
object this vignette assumes throughout. *Mapping optima and the orchestra
manifest* covers `goal = "map"`, a different axis of the same selector,
including what happens when no installed engine can honour the request.
*Combinatorial optimisation and delta evaluation* covers the one design
space `auto` never races: permutations, routed instead to a dedicated
annealer.

## Reproduce

Seed `42` for the noisy-objective example, `1` for the race, `2` for the
explicit race, `3` for the budget cap, and `5` for every run in the budget
sweep.

```{r session-info}
sessionInfo()
```
