---
title: "Combinatorial optimisation and delta evaluation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Combinatorial optimisation and delta evaluation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(optimix)
```

## Orderings as a design space

Not every decision variable is a real number. Scheduling, routing, and
assignment problems search over orderings, and `space_permutation(n)`
declares exactly that: a design space whose points are the permutations of
`seq_len(n)`. It is used through the power path -- build an `optim_problem()`
whose objective takes an integer permutation and returns a single number.

The engine behind it is `perm_sa`, a swap-move simulated annealer with
Metropolis acceptance and geometric cooling. It needs nothing beyond base R,
so everything on this page runs on a machine with no suggested packages
installed, and `method = "auto"` routes every permutation problem to it. It
is a general-purpose permutation engine, not a specialist: on large instances
of a structured problem such as the travelling salesman, a dedicated solver
with problem-specific moves will beat it.

## A worked example

Twenty items of different weights must be placed into twenty positions of
increasing cost, paying the item's weight times the position's cost. The best
ordering puts the heaviest items in the cheapest positions:

```{r data}
n <- 20L
set.seed(101)
w <- as.numeric(sample.int(100L, n))
pos_cost <- as.numeric(seq_len(n))
placement_cost <- function(perm) {
  sum(w[perm] * pos_cost)
}
```

## The delta hook

Simulated annealing proposes thousands of swaps, and re-evaluating the whole
objective after each one is wasteful when a swap only touches a sliver of the
sum. The contract's `delta_fn` hook makes that saving first-class. Its
signature is `delta_fn(perm, i, j)`, and it must return the change in the
objective when the entries at positions `i` and `j` of `perm` are swapped.
For the placement objective the change involves only the two moved items:

```{r delta}
placement_delta <- function(perm, i, j) {
  (w[perm[i]] - w[perm[j]]) * (pos_cost[j] - pos_cost[i])
}
```

A wrong delta would corrupt the anneal silently, so optimix does not take the
hook on trust. At startup the claimed delta of one random swap is verified
against a full re-evaluation and a disagreement stops with an error. During
the anneal the accumulated value is re-synchronised periodically with a fresh
full evaluation, and the value reported in the result is always a fresh full
evaluation at the best permutation found -- never the accumulated shortcut.

## Same trajectory, fewer evaluations

The two problems below are identical except for the hook, and they share a
seed. The engine draws its random numbers in the same order whether or not a
`delta_fn` is present, and the delta here is exact, so both runs walk the
same trajectory and return the same permutation:

```{r runs}
prob_plain <- optim_problem(
  fn = placement_cost,
  space = space_permutation(n),
  max_evals = 4000,
  seed = 7
)
prob_delta <- optim_problem(
  fn = placement_cost,
  space = space_permutation(n),
  max_evals = 4000,
  seed = 7,
  delta_fn = placement_delta
)
plain <- optimix(prob_plain)
fast <- optimix(prob_delta)
identical(plain$par, fast$par)
c(plain = plain$value, fast = fast$value)
```

In this seeded run the anneal lands on the exact optimum, which for this
objective can be computed directly:

```{r exact}
sum(sort(w, decreasing = TRUE) * pos_cost)
```

## Counting honestly

`counts[["function"]]` counts true objective evaluations only. Calls to the
`delta_fn` never touch the objective, so they are tallied separately in the
diagnostics field `delta_evals` rather than folded into the count:

```{r counts}
c(plain = plain$counts[["function"]], fast = fast$counts[["function"]])
c(plain = plain$diagnostics$delta_evals, fast = fast$diagnostics$delta_evals)
```

The plain run pays one full evaluation per scored move. With the hook, the
objective is touched only for the initial score, the startup contract check,
the periodic re-syncs, and the final fresh evaluation of the returned value
-- a few true evaluations in place of four thousand, for the same answer.
For objectives where a full evaluation is genuinely expensive (a delta over
a large assignment or schedule), this is the difference between an anneal
that explores and one that stalls.
