---
title: "Combinatorial optimisation and delta evaluation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Combinatorial optimisation and delta evaluation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~1 s under rmarkdown::render(); needs nothing beyond base
     R for the annealer itself, and ggplot2 for the trajectory figure.
     macOS arm64, R 4.5.1, one core. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150
)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r library}
library(optimix)
```

## Why

A logistics planner sequencing twenty tasks into twenty slots of rising
cost wants the cheapest ordering, and a simulated anneal needs thousands of
trial swaps to find it. Re-scoring the whole cost function from scratch
after every one of those swaps is wasteful when a single swap only touches
a sliver of the total -- and on a genuinely expensive objective, it is slow
enough to turn a search that would otherwise explore freely into one that
stalls. This vignette asks: can the search be told, cheaply, what changes
when only two items swap places, and still land on exactly the same
answer it would have found the slow way?

## What

`space_permutation(n)` declares a design space whose points are the
permutations of `seq_len(n)`, reached through the power path: build an
`optim_problem()` whose objective takes an integer permutation and returns
a single number. The engine behind it is `perm_sa`, a swap-move simulated
annealer with Metropolis acceptance and geometric cooling, needing nothing
beyond base R; `method = "auto"` routes every permutation problem to it.
The contract's `delta_fn` hook -- signature `delta_fn(perm, i, j)` -- lets
a problem declare the change in the objective from swapping positions `i`
and `j`, so the annealer can score a move without a full re-evaluation.

## Do

### Orderings as a design space

Not every decision variable is a real number. Scheduling, routing, and
assignment problems search over orderings, and `space_permutation(n)`
declares exactly that. It is a general-purpose permutation engine, not a
specialist: on large instances of a structured problem such as the
travelling salesman, a dedicated solver with problem-specific moves will
beat it.

### A worked example

A synthetic assignment problem: twenty items of different weights must be
placed into twenty positions of increasing cost, paying the item's weight
times the position's cost. The best ordering puts the heaviest items in
the cheapest positions:

```{r data}
n <- 20L
set.seed(101)
w <- as.numeric(sample.int(100L, n))
pos_cost <- as.numeric(seq_len(n))
placement_cost <- function(perm) {
  sum(w[perm] * pos_cost)
}
```

### The delta hook

Simulated annealing proposes thousands of swaps, and re-evaluating the
whole objective after each one is wasteful when a swap only touches a
sliver of the sum. The contract's `delta_fn` hook makes that saving
first-class. Its signature is `delta_fn(perm, i, j)`, and it must return
the change in the objective when the entries at positions `i` and `j` of
`perm` are swapped. For the placement objective the change involves only
the two moved items:

```{r delta}
placement_delta <- function(perm, i, j) {
  (w[perm[i]] - w[perm[j]]) * (pos_cost[j] - pos_cost[i])
}
```

A wrong delta would corrupt the anneal silently, so optimix does not take
the hook on trust. At startup the claimed delta of one random swap is
verified against a full re-evaluation and a disagreement stops with an
error. During the anneal the accumulated value is re-synchronised
periodically with a fresh full evaluation, and the value reported in the
result is always a fresh full evaluation at the best permutation found --
never the accumulated shortcut.

### Same trajectory, fewer evaluations

The two problems below are identical except for the hook, and they share a
seed. To see not just how many true evaluations each run spends but where
they land, `placement_cost` is wrapped in a small counter that logs every
call it actually receives, without changing what it returns:

```{r tracker}
track_calls <- function(fn) {
  log_env <- new.env(parent = emptyenv())
  log_env$vals <- numeric(0)
  wrapped <- function(x) {
    v <- fn(x)
    log_env$vals[length(log_env$vals) + 1L] <- v
    v
  }
  list(fn = wrapped, log = log_env)
}
```

The engine draws its random numbers in the same order whether or not a
`delta_fn` is present, and the delta here is exact, so both runs walk the
same trajectory and return the same permutation:

```{r runs}
tr_plain <- track_calls(placement_cost)
prob_plain <- optim_problem(
  fn = tr_plain$fn,
  space = space_permutation(n),
  max_evals = 4000,
  seed = 7
)
tr_delta <- track_calls(placement_cost)
prob_delta <- optim_problem(
  fn = tr_delta$fn,
  space = space_permutation(n),
  max_evals = 4000,
  seed = 7,
  delta_fn = placement_delta
)
plain <- optimix(prob_plain)
fast <- optimix(prob_delta)
identical(plain$par, fast$par)
c(plain = plain$value, fast = fast$value)
```

In this seeded run the anneal lands on the exact optimum, which for this
objective can be computed directly:

```{r exact}
exact_optimum <- sum(sort(w, decreasing = TRUE) * pos_cost)
exact_optimum
```

### Counting true evaluations

`counts[["function"]]` counts true objective evaluations only. Calls to
the `delta_fn` never touch the objective, so they are tallied separately
in the diagnostics field `delta_evals` rather than folded into the count:

```{r counts}
c(plain = plain$counts[["function"]], fast = fast$counts[["function"]])
c(plain = plain$diagnostics$delta_evals, fast = fast$diagnostics$delta_evals)
```

### A figure: the sparse footprint of true evaluations

The counts above say how many true evaluations each run spent; they do not
show what those evaluations achieved along the way. The tracker attached
above recorded every value `placement_cost` actually returned, in call
order, for both runs -- `r length(tr_plain$log$vals)` calls for the plain
run and `r length(tr_delta$log$vals)` for the delta-hook run. Plotting the
running best (the cumulative minimum) against call order shows the shape
of the search each spent its evaluations on:

```{r trace-kable, echo = FALSE}
trace_summary <- data.frame(
  run = c("plain", "delta_fn attached"),
  true_evaluations = c(length(tr_plain$log$vals), length(tr_delta$log$vals)),
  final_best = c(min(tr_plain$log$vals), min(tr_delta$log$vals))
)
knitr::kable(
  trace_summary, digits = 0, row.names = FALSE,
  caption = "True evaluations spent and the best cost each run reached."
)
```

```{r fig-trace, echo = has_ggplot2, eval = has_ggplot2, fig.cap = "Best placement cost found so far, in call order, for the plain run against the delta_fn run. The plain run's true evaluations trace a dense staircase over four thousand calls; the delta_fn run reaches the same floor from a handful of calls, because most of its moves are scored from the delta alone and never touch the objective. Each panel keeps its own x-axis, since the two runs spend true evaluations at entirely different rates."}
trace_df <- rbind(
  data.frame(
    run = "plain (no delta_fn)", call_index = seq_along(tr_plain$log$vals),
    best_so_far = cummin(tr_plain$log$vals)
  ),
  data.frame(
    run = "delta_fn attached", call_index = seq_along(tr_delta$log$vals),
    best_so_far = cummin(tr_delta$log$vals)
  )
)
trace_df$run <- factor(
  trace_df$run, levels = c("plain (no delta_fn)", "delta_fn attached")
)
ggplot2::ggplot(
  trace_df, ggplot2::aes(x = call_index, y = best_so_far, colour = run)
) +
  ggplot2::geom_step() +
  ggplot2::geom_point(size = 1.6) +
  ggplot2::facet_wrap(~run, scales = "free_x") +
  ggplot2::scale_colour_viridis_d(name = "run", guide = "none") +
  ggplot2::labs(
    x = "true objective evaluations, in call order",
    y = "best placement cost found so far (synthetic units)",
    title = "The delta hook reaches the same floor from a handful of calls"
  ) +
  ggplot2::theme_minimal()
```

```{r fig-trace-skip, echo = FALSE, eval = !has_ggplot2, results = "asis"}
cat("ggplot2 is not installed; the trajectory figure is skipped.\n")
```

The plain run pays one full evaluation per scored move. With the hook, the
objective is touched only for the initial score, the startup contract
check, the periodic re-syncs, and the final fresh evaluation of the
returned value -- a few true evaluations in place of four thousand, for
the same answer. For objectives where a full evaluation is genuinely
expensive (a delta over a large assignment or schedule), this is the
difference between an anneal that explores and one that stalls.

## Read

Both runs return the identical permutation
(`r identical(plain$par, fast$par)`), at a value of `r plain$value`, which
is exactly the value a direct calculation confirms is optimal for this
instance: `r exact_optimum`. The plain run spent
`r plain$counts[["function"]]` true evaluations against a `max_evals` of
4000; the delta-hook run spent `r fast$counts[["function"]]`, with
`r fast$diagnostics$delta_evals` further calls to `delta_fn` that never
touched the objective at all. The trajectory table and figure make that
gap concrete rather than abstract: the plain run's staircase runs the
whole width of its panel, improving in small steps across thousands of
calls, while the delta-hook panel needs only
`r length(tr_delta$log$vals)` points to land on the same floor of
`r min(tr_delta$log$vals)`. Neither panel's y-axis is a different scale --
both bottom out at the same value -- which is the visual confirmation that
fewer evaluations did not mean a worse answer here.

## Limits

This instance is small enough (twenty items) that the plain run's four
thousand evaluations cost nothing worth economising in absolute terms; the
delta hook earns its keep on objectives where a single full evaluation is
itself expensive, which this synthetic placement cost is not. The delta
here is also exact, verified once at startup and re-synchronised
periodically during the anneal; a `delta_fn` that drifts from the true
objective between re-syncs, rather than disagreeing outright, is not a
failure mode this vignette exercises. `perm_sa` is a general-purpose
annealer, and this vignette does not compare it against a specialist
solver for a structured combinatorial problem such as the travelling
salesman, where a dedicated method is expected to win.

## What to read next

*Getting started with optimix* covers the two front doors and the result
object this vignette builds on. *The auto selector and racing* explains
why `method = "auto"` never races a permutation space -- it routes
directly to `perm_sa` instead. *Mapping optima and the orchestra manifest*
covers the continuous-space counterpart to "one best point": the full set
of equally good optima, which is not defined over an ordering.

## Reproduce

Seed `101` for the synthetic weight vector; seed `7` for both anneal runs.

```{r session-info}
sessionInfo()
```
