---
title: "Mapping optima and the orchestra manifest"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mapping optima and the orchestra manifest}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(optimix)
```

## When one argmin is not the question

An optimiser usually answers "what is the single best point". On a multimodal
problem that answer hides information: a design task may have several
solutions of essentially equal quality, and which one is preferable can
depend on constraints the objective never saw. The useful answer is then the
whole set of good solutions, and `optimix_map()` returns exactly that.

The classic demonstration is Himmelblau's function, a two-dimensional surface
with four global minima of equal depth:

```{r himmelblau-def}
himmelblau <- function(x) {
  (x[1]^2 + x[2] - 11)^2 + (x[1] + x[2]^2 - 7)^2
}
```

## Two paths, one function

`optimix_map()` has a native path and a premium path, and picks between them
automatically. The native path is always available: it runs a local search
from many Latin-hypercube starts and clusters the converged points into
distinct optima, using nothing beyond base R. When a new-enough proxymix is
installed, the premium path returns a calibrated Gaussian-mixture map over
the optima instead -- a queryable posterior, carried in the result's `map`
field. The result's `source` field reports which path produced it:

```{r map}
set.seed(11)
map <- optimix_map(himmelblau, lower = c(-6, -6), upper = c(6, 6))
map
map$source
```

The two paths answer the same question in different registers. On the native
path each optimum is polished by a local search, so the four printed values
sit at effectively zero and the locations are sharp. On the proxymix path the
modes are peaks of a fitted mixture -- a summary of where the low-value mass
lies -- so locations are approximate and a shallow extra mode can appear
beside the four true wells. The modes are printed best first in either case,
so deep optima separate from shallow artefacts at a glance, and the `modes`
and `values` fields give the full set programmatically.

## Asking auto for a map

The same intent can be declared on the main verb. `goal = "map"` steers
`method = "auto"` towards a map-emitting engine, and the result carries the
queryable mixture in `$map`:

```{r goal-map, eval = requireNamespace("proxymix", quietly = TRUE)}
set.seed(12)
res <- optimix(himmelblau, lower = c(-6, -6), upper = c(6, 6), goal = "map")
summary(res)
```

The fallback is honest. When no map-emitting engine fits the problem --
proxymix absent, or the dimension beyond its range -- `auto` does not fake a
map. It returns the single best optimum through the usual rules and says so
in the provenance, so a consumer can always tell what it received. Note also
the evaluation count above: the mixture path spends many cheap evaluations
to map the whole optimum set, so it suits inexpensive objectives, and the
selector reaches it only when a map is asked for.

## The orchestra manifest

An optimisation result often feeds a downstream system rather than a person,
and `as_orchestra_manifest()` re-expresses a result as a typed hand-off
contract for that purpose. The optimum rides in `params` as a one-row table.
The typed summary and metadata carry the engine, the convergence verdict,
and the true evaluation count. The `uncertainty_available` flag records
whether the manifest carries a posterior over the optimum (a mixture map) or
a bare point estimate, so a consumer never mistakes one for the other. The
payload is integrity-hashed at emission, and `verify_manifest()` recomputes
the hash to detect any later modification:

```{r manifest}
best <- optimix(himmelblau, lower = c(-6, -6), upper = c(6, 6),
                method = "base_optim")
m <- as_orchestra_manifest(best)
m@params
m@summary$headline
m@metadata$uncertainty_available
verify_manifest(m)
```

Here a single-point engine produced the result, so the flag is `FALSE` and
the manifest honestly offers a point estimate only. Had the proxymix mixture
engine run, the flag would be `TRUE` and the metadata would carry the full
solution map alongside the optimum.
