---
title: "Mapping optima and the orchestra manifest"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mapping optima and the orchestra manifest}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~1 s under rmarkdown::render() with proxymix installed
     (the premium mixture path is fast on a two-dimensional problem); a
     build without proxymix still renders every section, falling back to
     the native multi-start path and the documented goal-abstention text.
     macOS arm64, R 4.5.1, one core. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5.5,
  dpi = 150
)
has_proxymix <- requireNamespace("proxymix", quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r library}
library(optimix)
```

## Why

A breeder ranking candidate sowing configurations against a cost function
does not always want the single best combination back. If several
configurations score within noise of each other, the one to actually plant
can depend on constraints the cost function never saw -- seed availability,
rotation history, machinery on hand -- and a plain argmin search shows only
the first of those configurations it happened to converge to, never the
rest. The question this vignette answers is: *when a design surface has
several solutions of essentially equal merit, can optimix hand back all of
them, state plainly when it cannot, and package the answer for a system
downstream rather than a person reading a console?*

## What

`optimix_map()` returns the distinct optima of a function over a box,
rather than a single best point. It has a native path (always available,
local search from many starts, clustered) and a premium path (a calibrated
Gaussian-mixture map, when a new-enough proxymix is installed), and reports
which one ran in its `source` field. The same intent reaches the main verb
through `goal = "map"` on `optimix()`: `auto` then routes to a map-emitting
engine, or declines and says so plainly when none fits, marking the returned
`optimix_result` with the `optimix_map_abstention` class so a caller can
detect the decline without parsing prose. `as_orchestra_manifest()`
re-expresses any result -- mapped or not -- as an `orchestra_manifest`, the
federation's shared, hashed, typed hand-off object (contract C2 v2 in
`ORCHESTRA.md`'s registry); `verify_manifest()` checks that hash.

## Do

### A synthetic surface with four equally good answers

The classic demonstration is a synthetic test surface, Himmelblau's
function, a two-dimensional cost surface with four global minima of equal
depth:

```{r himmelblau-def}
himmelblau <- function(x) {
  (x[1]^2 + x[2] - 11)^2 + (x[1] + x[2]^2 - 7)^2
}
```

### Two paths, one function

`optimix_map()` has a native path and a premium path, and picks between
them automatically. The native path is always available: it runs a local
search from many Latin-hypercube starts and clusters the converged points
into distinct optima, using nothing beyond base R. When a new-enough
proxymix is installed, the premium path returns a calibrated
Gaussian-mixture map over the optima instead -- a queryable posterior,
carried in the result's `map` field. The result's `source` field reports
which path produced it:

```{r map}
set.seed(11)
map <- optimix_map(himmelblau, lower = c(-6, -6), upper = c(6, 6))
map
map$source
```

```{r map-kable, echo = FALSE}
modes_tab <- data.frame(
  rank = seq_len(map$n), x1 = map$modes[, 1L], x2 = map$modes[, 2L],
  value = map$values
)
knitr::kable(
  modes_tab, digits = 4, row.names = FALSE,
  caption = paste0(
    "Every optimum `optimix_map()` recovered (source: ", map$source, ")."
  )
)
```

```{r fig-map, echo = has_ggplot2, eval = has_ggplot2, fig.cap = "Himmelblau's synthetic function, capped at 500 to keep the four wells legible; the points found by optimix_map() are marked in red, best (deepest) first. Four sit in visually identical, fully dark wells; a fifth, shallower point, when the premium path finds one, is set apart by sitting in a lighter band -- exactly the shallow-artefact distinction a single printed value cannot show."}
gx <- seq(-6, 6, length.out = 120L)
gy <- seq(-6, 6, length.out = 120L)
grid_df <- expand.grid(x1 = gx, x2 = gy)
grid_df$z <- apply(
  grid_df, 1L, function(row) himmelblau(c(row[["x1"]], row[["x2"]]))
)
grid_df$z <- pmin(grid_df$z, 500)
modes_df <- data.frame(
  x1 = map$modes[, 1L], x2 = map$modes[, 2L], rank = seq_len(map$n)
)
ggplot2::ggplot(grid_df, ggplot2::aes(x1, x2, z = z)) +
  ggplot2::geom_contour_filled(bins = 14L) +
  ggplot2::scale_fill_viridis_d(name = "objective\n(capped at 500)",
                                guide = "none") +
  ggplot2::geom_point(
    data = modes_df, ggplot2::aes(x1, x2), inherit.aes = FALSE,
    colour = "white", fill = "firebrick", shape = 21L, size = 3L, stroke = 1L
  ) +
  ggplot2::labs(
    x = "x1", y = "x2",
    title = "Himmelblau's synthetic surface: every recovered optimum"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::coord_equal()
```

```{r fig-map-skip, echo = FALSE, eval = !has_ggplot2, results = "asis"}
cat("ggplot2 is not installed; the contour figure is skipped.\n")
```

The marked points confirm that `optimix_map()` returned a genuine set of
distinct optima rather than the same well found repeatedly -- the outcome
a plain argmin search cannot show, since it stops at the first optimum it
converges to.

The two paths answer the same question in different registers. On the
native path each optimum is polished by a local search, so the printed
values sit at effectively zero and the locations are sharp. On the
proxymix path the modes are peaks of a fitted mixture -- a summary of
where the low-value mass lies -- so locations are approximate and a
shallow extra mode can appear beside the true wells. The modes are printed
best first in either case, so deep optima separate from shallow artefacts
at a glance, and the `modes` and `values` fields give the full set
programmatically.

### Asking auto for a map

The same intent can be declared on the main verb. `goal = "map"` steers
`method = "auto"` towards a map-emitting engine, and the result carries the
queryable mixture in `$map`:

```{r goal-map, eval = has_proxymix, echo = has_proxymix}
set.seed(12)
res <- optimix(himmelblau, lower = c(-6, -6), upper = c(6, 6), goal = "map")
summary(res)
```

```{r goal-map-skip, eval = !has_proxymix, echo = FALSE, results = "asis"}
cat("proxymix is not installed on this build, so this chunk is not run;",
    "the fallback described below is what auto does in exactly this",
    "situation.\n")
```

The mixture path itself spends many cheap evaluations to map the whole
optimum set -- `res$counts` runs well above a single-point search's budget
when proxymix is installed and the chunk above executes -- so it suits
inexpensive objectives, and the selector reaches it only when a map is
asked for.

### When the map cannot be honoured

The fallback does not fabricate a map, and it is worth seeing happen
rather than taking on trust. `proxymix_map` is registered for problems of dimension 10 or
fewer; a 12-dimensional request for `goal = "map"` is therefore outside
its range whether or not proxymix is installed, so `auto` cannot honour it
and falls back to the single best optimum instead of fabricating a map:

```{r abstention}
set.seed(7)
res_abstain <- optimix(
  function(x) sum((x - 0.4)^2), lower = rep(-3, 12L), upper = rep(3, 12L),
  goal = "map"
)
class(res_abstain)
inherits(res_abstain, "optimix_map_abstention")
res_abstain$provenance$engine
res_abstain$provenance$why
```

`optimix_map_abstention` is prepended to the result's class vector rather
than replacing it, so every ordinary `optimix_result` method still applies,
and a caller who only wants to know whether the request was honoured can
check `inherits(x, "optimix_map_abstention")` without parsing the `why`
string. The same prepend-not-replace convention marks a non-converged
optimum's manifest with `optimix_manifest_abstention`, so both declines are
recognisable from the class vector alone.

### The orchestra manifest

An optimisation result often feeds a downstream system rather than a
person, and `as_orchestra_manifest()` re-expresses a result as a typed
hand-off contract for that purpose -- the C2 v2 generic `orchestra_manifest`
in `ORCHESTRA.md`'s contract registry. The optimum rides in `params` as a
one-row table. The typed summary and metadata carry the engine, the
convergence verdict, and the true evaluation count. The
`uncertainty_available` flag records whether the manifest carries a
posterior over the optimum (a mixture map) or a bare point estimate, so a
consumer never mistakes one for the other. The payload is
integrity-hashed at emission, and `verify_manifest()` recomputes the hash
to detect any later modification:

```{r manifest}
best <- optimix(himmelblau, lower = c(-6, -6), upper = c(6, 6),
                method = "base_optim")
m <- as_orchestra_manifest(best)
m@params
m@summary$headline
m@metadata$uncertainty_available
verify_manifest(m)
```

Here a single-point engine produced the result, so the flag is `FALSE` and
the manifest offers only a point estimate, with no invented uncertainty.
Had the proxymix
mixture engine run, the flag would be `TRUE` and the metadata would carry
the full solution map alongside the optimum.

## Read

```{r read-prep, echo = FALSE}
class_str <- paste(class(res_abstain), collapse = ", ")
proxymix_para <- if (has_proxymix) {
  paste0(
    "With proxymix installed, asking `auto` for a map directly reaches `",
    res$provenance$engine, "` and spends ", res$counts[["function"]],
    " evaluations mapping the whole optimum set, an order of magnitude ",
    "above a single-point search's typical budget -- the price paid for ",
    "a posterior over every optimum rather than one point."
  )
} else {
  paste0(
    "proxymix is not installed on this build, so the direct ",
    "`goal = \"map\"` request above did not run; the fallback text ",
    "below is what replaced it either way, on a problem outside the ",
    "mixture engine's range."
  )
}
```

```{r read-text, echo = FALSE, results = "asis"}
cat(
  "Native or premium, `optimix_map()` recovered ", map$n, " optima on ",
  "this run, reported through the result's `source` field as \"",
  map$source, "\". The table lists every one of them; the figure shows ",
  "why the count alone is not the full picture -- whichever points sit ",
  "in a genuinely dark well are the deep, essentially tied optima, and ",
  "any point in a lighter band beside them is a shallow artefact of the ",
  "fitting procedure rather than a fifth true minimum.\n\n",
  sep = ""
)
cat(proxymix_para, "\n\n", sep = "")
cat(
  "The 12-dimensional request makes the fallback concrete rather than ",
  "descriptive: its class carries ", class_str, ", confirming ",
  "`optimix_map_abstention` sits first, and its provenance names the ",
  "engine that actually ran (`", res_abstain$provenance$engine,
  "`) and the reason the map itself was declined. The manifest section ",
  "closes the loop from either path to a downstream consumer: ",
  "`m@summary$headline` reads \"", m@summary$headline, "\", ",
  "`verify_manifest(m)$ok` is ", verify_manifest(m)$ok, ", and ",
  "`metadata$uncertainty_available` is ", m@metadata$uncertainty_available,
  " because `base_optim` produced this particular manifest -- a point, ",
  "correctly labelled as one.\n",
  sep = ""
)
```

A package-level routing study compared optimix's engines against
proxymix's `from_objective()` mapper across seven case-classes on
functions with known analytic optima, independent of both packages
(`comparisons/from_objective/`, ORCHESTRA_dev, 2026-06-24); its tally is
cited here and not recomputed. optimix won five of the seven classes --
every regime with a single best point to find, smooth or multimodal,
low-dimensional or high -- and proxymix won two: recovering every optimum
of a multimodal surface in one fit, and returning a genuine posterior over
the optimum set. That is exactly the split this vignette walks: reach for
`optimix_map()` when the deliverable is the whole set with a queryable
map, and for plain `optimix()` when it is the single best point.

## Limits

Himmelblau's function is a synthetic two-dimensional textbook surface with
four exactly equal minima; it is chosen because the ground truth is known,
not because it resembles a fielded design problem, and nothing here
demonstrates how many equally good optima a real cost surface tends to
carry. The premium path's mode locations are read off a fitted mixture, not
polished by a further local search, so their precision is lower than the
native path's; this vignette does not quantify that gap numerically; a
reader who needs an error bound on a proxymix-reported location should
treat it as approximate. The 12-dimensional abstention example is chosen
to guarantee the decline deterministically, regardless of whether proxymix
is installed; it does not explore the boundary case of a problem just
inside the mixture engine's dimension range, where the map may still
degrade gracefully rather than abstain outright.

## What to read next

*Getting started with optimix* covers the two front doors and the result
object this vignette builds on. *The auto selector and racing* covers the
rest of the selector's decision surface -- noise, expense, budget -- that
`goal = "map"` sits alongside. *Combinatorial optimisation and delta
evaluation* covers the one design space neither mapping nor racing
reaches: permutations.

## Reproduce

Seed `11` for the map, `12` for the `goal = "map"` call, `7` for the
12-dimensional abstention example.

```{r session-info}
sessionInfo()
```
