---
title: "Getting started with optimix"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with optimix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(optimix)
```

## One verb

`optimix` exposes a single verb. The easy path takes a function and box
bounds, in the spirit of `stats::optim()`:

```{r easy}
result <- optimix(
  function(x) sum(x^2),
  lower = c(-5, -5),
  upper = c(5, 5),
  method = "base_optim"
)
result
```

`minimise()` and `maximise()` are readable aliases for the same call.

## Choosing the optimiser

With the default `method = "auto"`, optimix selects an optimiser for the
problem from those whose package is installed, and records the choice in the
result's provenance. `list_optimisers()` shows what is available:

```{r registry}
list_optimisers()
```

## An `optim()`-compatible result

The result is shaped like a `stats::optim()` return value -- `par`, `value`,
`counts`, `convergence`, `message` -- so existing code keeps working, with
optimix extras alongside. The usual accessors apply:

```{r accessors}
coef(result)
as.data.frame(result)
as_optim(result)
```

## The power path

For noise, an evaluation budget, a warm start, or a seed, build an
`optim_problem()` directly and pass it to `optimix()`:

```{r power}
problem <- optim_problem(
  fn = function(x) sum((x - 1)^2),
  space = space_box(lower = c(-5, -5), upper = c(5, 5)),
  max_evals = 2000
)
optimix(problem, method = "base_optim")
```

The same problem object is what the easy path builds for you behind the
scenes -- one contract, two front doors.
