---
title: "Getting started with optimix"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with optimix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~1 s under rmarkdown::render() with cmaes, DEoptim,
     DEoptimR, dfoptim, DiceKriging, GenSA, nloptr and proxymix all
     installed; a build with none of those still renders every chunk here,
     because this vignette never needs an optional engine by name. macOS
     arm64, R 4.5.1, one core. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150
)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r library}
library(optimix)
```

## Why

An agronomist has written a cost function over two decision variables --
say a nitrogen rate and a sowing density -- and wants the combination that
minimises it under a fixed number of trial runs. She does not know, before
she starts, whether that cost surface is a smooth bowl, has a ripple hiding
a second dip, or is something else entirely, and she does not want to
choose a different optimisation package for every shape she might meet.
`optimix` is built for exactly that: one call that takes a function and a
box, and returns the best point it can find within the budget, whatever the
surface turns out to be. This vignette walks the whole surface of the
package -- both front doors, the result it hands back, and what the
optimiser actually found -- and closes by showing that optimum sitting
inside the landscape it was found in.

## What

`optimix()` is the single verb: it takes an objective function (or an
`optim_problem()`) and a design space, chooses or uses a named optimiser,
and returns an `optimix_result`. `optim_problem()` is the typed contract
behind it -- fn, `space` (typically a `space_box()`), an optional evaluation
budget, warm start and seed -- and is what the easy path builds internally
from `fn`, `lower` and `upper`. `list_optimisers()` reports the registry of
engines optimix knows about, which of their backend packages are installed,
and the traits (`global`, `noise_tolerant`, `emits_map`) the selector reads.
The result behaves like a `stats::optim()` return value, with `coef()`,
`as.data.frame()` and `as_optim()` as read-outs for code that expects
exactly that shape.

## Do

### The easy path

The easy path takes a function and box bounds, in the spirit of
`stats::optim()`. The objective below is a synthetic quadratic bowl centred
on the origin, and `method = "base_optim"` names the zero-dependency core
engine directly, so this chunk needs nothing beyond base R:

```{r easy}
result <- optimix(
  function(x) sum(x^2),
  lower = c(-5, -5),
  upper = c(5, 5),
  method = "base_optim"
)
result
```

`minimise()` and `maximise()` are readable aliases for the same call.

### Which optimisers are available

With the default `method = "auto"`, optimix selects an optimiser for the
problem from those whose package is installed, and records the choice in
the result's provenance. `list_optimisers()` shows what is available on
this build:

```{r registry}
lo <- list_optimisers()
lo
```

```{r registry-kable, echo = FALSE}
knitr::kable(
  lo, row.names = FALSE,
  caption = paste0(
    "The optimix engine registry on this build (", sum(lo$installed),
    " of ", nrow(lo), " engines installed)."
  )
)
```

### An `optim()`-compatible result

The result is shaped like a `stats::optim()` return value -- `par`,
`value`, `counts`, `convergence`, `message` -- so existing code keeps
working, with optimix extras alongside. The usual accessors apply:

```{r accessors}
coef(result)
as.data.frame(result)
as_optim(result)
```

### The power path

For noise, an evaluation budget, a warm start, or a seed, build an
`optim_problem()` directly and pass it to `optimix()`. The objective below
is a synthetic bowl shifted to `(1, 1)`, capped at 2000 evaluations:

```{r power}
problem <- optim_problem(
  fn = function(x) sum((x - 1)^2),
  space = space_box(lower = c(-5, -5), upper = c(5, 5)),
  max_evals = 2000
)
power_result <- optimix(problem, method = "base_optim")
power_result
```

The same problem object is what the easy path builds behind the scenes --
one contract, two front doors.

### A figure: where the optimum sits in the landscape

None of the results above shows what the objective actually looks like
around the point that was returned. The synthetic surface below is a bowl
centred near `(2, -1.5)` with a small sinusoidal ripple layered on top, so
it is not quite the clean quadratic of the two examples above; `method =
"auto"` is left to choose the engine, exactly as a first-time user would
call it:

```{r auto-optimum}
set.seed(9)
rippled_bowl <- function(x) {
  sum((x - c(2, -1.5))^2) + 0.4 * sin(3 * x[1]) * cos(2 * x[2])
}
auto_result <- optimix(
  rippled_bowl, lower = c(-5, -5), upper = c(5, 5), method = "auto"
)
auto_result
auto_result$provenance$engine
```

```{r fig-landscape, echo = has_ggplot2, eval = has_ggplot2, fig.cap = "The rippled synthetic bowl over its full search box, with the optimum optimix found marked in red. The ripple bends the contours away from a perfect ellipse without moving the deepest point far from the unperturbed bowl's centre."}
gx <- seq(-5, 5, length.out = 100L)
gy <- seq(-5, 5, length.out = 100L)
grid_df <- expand.grid(x1 = gx, x2 = gy)
grid_df$z <- apply(
  grid_df, 1L, function(row) rippled_bowl(c(row[["x1"]], row[["x2"]]))
)
opt_df <- data.frame(x1 = auto_result$par[1L], x2 = auto_result$par[2L])
ggplot2::ggplot(grid_df, ggplot2::aes(x1, x2, z = z)) +
  ggplot2::geom_contour_filled(bins = 12L) +
  ggplot2::scale_fill_viridis_d(name = "cost", guide = "none") +
  ggplot2::geom_point(
    data = opt_df, ggplot2::aes(x1, x2), inherit.aes = FALSE,
    colour = "white", fill = "firebrick", shape = 21L, size = 3L, stroke = 1L
  ) +
  ggplot2::labs(
    x = "x1 (nitrogen-rate-style decision variable)",
    y = "x2 (sowing-density-style decision variable)",
    title = "Where the returned optimum sits in the full landscape"
  ) +
  ggplot2::theme_minimal()
```

```{r fig-landscape-skip, echo = FALSE, eval = !has_ggplot2, results = "asis"}
cat("ggplot2 is not installed; the landscape figure is skipped.\n")
```

## Read

The easy-path bowl is solved exactly: `optimix()` returns a value of
`r round(result$value, 6)` at par = (`r paste(result$par, collapse = ", ")`),
using `r result$counts[["function"]]` evaluations of the zero-dependency
`base_optim` engine. The registry carries `r nrow(lo)` named optimisers, of
which `r sum(lo$installed)` are installed and available on this build --
every engine behind a `Suggests` package plus the base engine and the
permutation annealer, since every optional package checked in the DESCRIPTION
happens to be present here. The accessors read the same result three ways:
`coef()` returns the bare parameter vector, `as.data.frame()` adds the
objective value and the engine name as columns, and `as_optim()` strips the
result back down to the five components a `stats::optim()`-based caller
expects. The power-path bowl, shifted to `(1, 1)`, is also solved exactly,
in `r power_result$counts[["function"]]` evaluations against its 2000-call
budget -- the budget is a ceiling the engine did not need to spend.

The rippled bowl is where a single number stops being enough. `method =
"auto"` chose `r auto_result$provenance$engine` for it, reaching a value of
`r round(auto_result$value, 4)` at
par = (`r paste(round(auto_result$par, 3), collapse = ", ")`).
Read on its own, that pair of numbers says nothing about how close the
ripple came to hiding a second, competing dip nearby, or how far the
optimum sits from the edges of the search box. The figure answers both at
a glance: the returned point sits inside the deepest contour band, close to
the unperturbed bowl's centre at `(2, -1.5)`, and the ripple bends the bands
without creating a second dark region the optimum could have missed.

## Limits

This vignette runs one call per example and reads its result in isolation;
it does not compare engines against each other, examine why `method =
"auto"` chose one engine over another, or characterise a budget's effect on
solution quality -- *The auto selector and racing* does both. It also stays
in two dimensions, in a continuous box, with a single global optimum in
each example: `optimix_map()` for a surface with several equally good
optima, and `space_permutation()` for search over orderings rather than
real-valued points, are both left to their own vignettes. Nothing here
demonstrates that a chosen engine failed to converge, or shows the classed
abstention a caller can check for when it did; that machinery is exercised
directly in *Mapping optima and the orchestra manifest*.

## What to read next

*The auto selector and racing* explains the rules and the race behind
`method = "auto"`, states the held-out evidence for when it wins, and
introduces the provenance fields this vignette only prints in passing.
*Mapping optima and the orchestra manifest* answers "what if there is more
than one good answer", and shows the typed abstention a caller receives
when a request cannot be honoured. *Combinatorial optimisation and delta
evaluation* moves from real-valued boxes to orderings, for scheduling- and
assignment-shaped problems.

## Reproduce

Seed `9` for the rippled-bowl figure; the two quadratic-bowl examples are
deterministic and draw no random numbers.

```{r session-info}
sessionInfo()
```
